---
title: "Alpha-beta-labortype"
author: "Tien Do"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r include=FALSE}
library(phyloseq)
library(ape)
library(tidyverse)
library(picante)
library(ggsignif)
library(ggplot2)
library(vegan)
library("car")
```
#### Load in RData ####
```{r}
load("hiv_rare.RData")
load("hiv_final.RData")
```
## color theme: #d7191c
#fdae61
#ffffbf
#abd9e9
#2c7bb6
#### Filter metadata ####
```{r}
#duplicate phylosesq object
phyloseq_hiv <- hiv_rare 
# Filter out 3rd visit samples:
phyloseq_filter_hiv <- subset_samples(phyloseq_hiv,Visit_Cat == "2nd Visit")
# change labor categories
labor_rename <- as.data.frame(sample_data(phyloseq_filter_hiv))
```
```{r}
labor_rename$Visit_Age <- ifelse(labor_rename$Visit_Age > 45, "Older than 45", "Younger than 45")
labor_rename$Labor_category <- ifelse(labor_rename$Labor_category %in% c("Non_Labor_Houseworker","Non_Labor_Other","Unemployed"), "Non-Labor", "Labor")
sample_data(phyloseq_filter_hiv) <- sample_data(labor_rename)

```
```{r}
#extract the sample data:
sample_phyloseq_hiv <- sample_data(phyloseq_filter_hiv)
#retain wanted columns:
phyloseq_labor<- sample_phyloseq_hiv[, c( "HIV_Status", "Labor_category","Education_Level","Visit_Age")]
# Replace the sample data in your phyloseq object with the updated one
sample_data(phyloseq_filter_hiv) <- phyloseq_labor
#alpha diversity metrics
alphadiv <- estimate_richness(phyloseq_filter_hiv)
samp_dat <- sample_data(phyloseq_filter_hiv)
samp_dat_wdiv <- data.frame(samp_dat, alphadiv)
# look for the total number of species in the subsetted data
total_species <- as.numeric(ntaxa(phyloseq_filter_hiv))
# calculate pielou's evenness index
samp_dat_wdiv$pielou <- samp_dat_wdiv$Shannon/log(total_species)
#replace back in phyloseq
sample_data(phyloseq_filter_hiv) <- samp_dat_wdiv

```
```{r}
#filter out 
hiv_pos <- subset_samples(phyloseq_filter_hiv, HIV_Status == "Positive")
hiv_pos_fr <- sample_data(hiv_pos)
hiv_pos_fr <- data.frame(hiv_pos_fr)
hiv_neg <- subset_samples(phyloseq_filter_hiv, HIV_Status =="Negative")
hiv_neg_fr <- sample_data(hiv_neg)
hiv_neg_fr <- data.frame(hiv_neg_fr)
hiv_labor <- subset_samples(phyloseq_filter_hiv, Labor_category == "Labor")
hiv_labor_fr <- sample_data(hiv_labor)
hiv_labor_fr <- data.frame(hiv_labor_fr)
hiv_nonlabor <- subset_samples(phyloseq_filter_hiv, Labor_category == "Non-Labor")
hiv_nonlabor_fr <- sample_data(hiv_nonlabor)
hiv_nonlabor_fr <- data.frame(hiv_nonlabor_fr)
#filter education level: tertiary, secondary, tertiary
hiv_primary <- subset_samples(phyloseq_filter_hiv, Education_Level == "Primary")
hiv_primary_fr <- sample_data(hiv_primary)
hiv_primary_fr <- data.frame(hiv_primary_fr)
hiv_secondary <- subset_samples(phyloseq_filter_hiv, Education_Level == "Secondary")
hiv_secondary_fr <- sample_data(hiv_secondary)
hiv_secondary_fr <- data.frame(hiv_secondary_fr)
hiv_tertiary <- subset_samples(phyloseq_filter_hiv, Education_Level == "Tertiary")
hiv_tertiary_fr <- sample_data(hiv_tertiary)
hiv_tertiary_fr <- data.frame(hiv_tertiary_fr)
hiv_young <- subset_samples(phyloseq_filter_hiv, Visit_Age == "Younger than 45")
hiv_young_fr <- sample_data(hiv_young)
hiv_young_fr <- data.frame(hiv_young_fr)
hiv_old <- subset_samples(phyloseq_filter_hiv, Visit_Age == "Older than 45")
hiv_old_fr <- sample_data(hiv_old)
hiv_old_fr <- data.frame(hiv_old_fr)
```


# Alpha diversity #

## Alpha Diversity: Shannon and Pielou's Diversity ##
```{r}
#run glm with Education and HIV Status as covariables
glm.pielou_ed <- glm(pielou ~ Education_Level + HIV_Status, data =samp_dat_wdiv,
              family = gaussian())
summary(glm.pielou_ed)
```
### graph pielou education level
```{r}
pielou_ed <- ggplot(samp_dat_wdiv, aes(x=Education_Level, y=pielou, fill=Education_Level)) +
  geom_boxplot() +
  geom_point(size = 2) +
  ylab("Pielou's Evenness index") +
  xlab('Education_Level') +
  theme_minimal()
pielou_ed
```
### graph pielou education of HIV-
```{r}
lm_ed_pie_neg <- lm(log(pielou) ~ Education_Level,data = hiv_neg_fr) 
anova_ed_pie_neg <- aov(lm_ed_pie_neg)
summary(anova_ed_pie_neg)
TukeyHSD(anova_ed_pie_neg)
plot.pie_neg_ed <- ggplot(sample_data(hiv_neg), aes(x=Education_Level, y=pielou,fill=Education_Level)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61","#ffffbf"))+
  ggtitle("Pielou Primary, Secondary, Tertiary of HIV-")+
  xlab("Education Level") +
  ylab("Eveness") +
  geom_signif(comparisons = list(c("Primary","Secondary"),c("Secondary","Tertiary"),c("Primary","Tertiary")),
              annotations = c("NS.","NS.","NS."), y_negition= c(0.72, 0.75, 0.79)) +
  theme_minimal()# plot any metadata category against the PD
plot.pie_neg_ed # view plot
ggsave(filename = "plot_pie_neg_ed.png"
       , plot.pie_neg_ed
       , height=4, width=6)
```
### graph pielou education of HIV+
```{r}
lm_ed_pie_pos <- lm(log(pielou) ~ Education_Level,data = hiv_pos_fr) 
anova_ed_pie_pos <- aov(lm_ed_pie_pos)
summary(anova_ed_pie_pos)
TukeyHSD(anova_ed_pie_pos)
plot.pie_pos_ed <- ggplot(sample_data(hiv_pos), aes(x=Education_Level, y=pielou,fill=Education_Level)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61","#ffffbf"))+
  ggtitle("Pielou Primary, Secondary, Tertiary of HIV+")+
  xlab("Education Level") +
  ylab("Eveness") +
  geom_signif(comparisons = list(c("Primary","Secondary"),c("Secondary","Tertiary"),c("Primary","Tertiary")),
              annotations = c("NS.","NS.","NS."), y_position= c(0.65, 0.68, 0.75)) +
  theme_minimal()# plot any metadata category against the PD
plot.pie_pos_ed # view plot
ggsave(filename = "plot_pie_pos_ed.png"
       , plot.pie_pos_ed
       , height=4, width=6)
```
### graph pielou labor cat of HIV+
```{r}
wilcox_pos_lab_pie <- wilcox.test(PD ~ Labor_category, data=hiv_pos_fr, exact=FALSE)
wilcox_pos_lab_pie #p=0.4078
plot.pie_pos_lab <- ggplot(sample_data(hiv_pos), aes(x=Labor_category, y=pielou,fill=Labor_category)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61","#ffffbf"))+
  ggtitle("Pielou Labor category of HIV+")+
  xlab("Labor category") +
  ylab("Eveness") +
  geom_signif(comparisons = list(c("Labor","Non-Labor")), map_signif_level= TRUE, test ="wilcox.test", y_position = 0.7) +
  theme_minimal()# plot any metadata category against the PD
plot.pie_pos_lab # view plot
ggsave(filename = "plot_pie_pos_lab.png"
       , plot.pie_pos_lab
       , height=4, width=6)
```
### graph pielou labor cat of HIV-
```{r}
plot.pie_neg_lab <- ggplot(sample_data(hiv_neg), aes(x=Labor_category, y=pielou,fill=Labor_category)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61","#ffffbf"))+
  ggtitle("Pielou Labor of HIV-")+
  xlab("labucation Level") +
  ylab("Eveness") +
  geom_signif(comparisons = list(c("Labor","Non-Labor")), map_signif_level= TRUE, test ="wilcox.test", y_negition = 0.7) +
  theme_minimal()# plot any metadata category against the PD
plot.pie_neg_lab # view plot
ggsave(filename = "plot_pie_neg_lab.png"
       , plot.pie_neg_lab
       , height=4, width=6)
```
### graph pielou age of HIV-
```{r}
plot.pie_neg_age <- ggplot(sample_data(hiv_neg), aes(x=Visit_Age, y=pielou,fill=Visit_Age)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61","#ffffbf"))+
  ggtitle("Pielou age of HIV-")+
  xlab("Age") +
  ylab("Eveness") +
  geom_signif(comparisons = list(c("Older than 45","Younger than 45")), map_signif_level= TRUE, test ="wilcox.test", y_negition = 0.7) +
  theme_minimal()# plot any metadata category against the PD
plot.pie_neg_age # view plot
ggsave(filename = "plot_pie_neg_age.png"
       , plot.pie_neg_age
       , height=4, width=6)
```
### Pielou Age of HIV+
```{r}
wilcox_pos_age_pie <- wilcox.test(pielou ~ Visit_Age, data=hiv_pos_fr, exact=FALSE)
wilcox_pos_age_pie #p=0.2387
plot.pie_pos_age <- ggplot(sample_data(hiv_pos), aes(x=Visit_Age, y=pielou,fill=Visit_Age)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61","#ffffbf"))+
  ggtitle("Pielou age of HIV+")+
  xlab("Age") +
  ylab("Eveness") +
  geom_signif(comparisons = list(c("Older than 45","Younger than 45")), map_signif_level= TRUE, test ="wilcox.test", y_position = 0.7) +
  theme_minimal()# plot any metadata category against the PD
plot.pie_pos_age # view plot
ggsave(filename = "plot_pie_pos_age.png"
       , plot.pie_pos_age
       , height=4, width=6)
```
## Richness ##
### Labor type within HIV-
```{r}

phylo_dist_neg <- pd(t(otu_table(hiv_neg)), phy_tree(hiv_neg),
                 include.root=F) # calculate Faith's Eveness as PD
sample_data(hiv_neg)$PD <- phylo_dist_neg$PD # add PD to metadata table
pd_neg <- sample_data(hiv_neg)$PD
hiv_neg_fr$PD <- pd_neg
wilcox_neg_labor <- wilcox.test(PD ~ Labor_category, data=hiv_neg_fr, exact=FALSE)$p.value #p=0.4078
plot.pd_neg <- ggplot(sample_data(hiv_neg), aes(Labor_category, PD,fill=Labor_category)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("Labor vs non-labor of HIV+")+
  xlab("Labor category") +
  ylab("Eveness") +
  geom_signif(comparisons = list(c("Labor","Non-Labor")),
              map_signif_level= TRUE, test ="wilcox.test", y_negition = 35)+ # plot any metadata category against the PD
  theme_minimal()

plot.pd_neg # view plot
ggsave(filename = "plot_PD_hivneg.png"
       , plot.pd_neg
       , height=4, width=6)
```
### Labor type within HIV+
```{r}
phylo_dist_pos <- pd(t(otu_table(hiv_pos)), phy_tree(hiv_pos),
                 include.root=F) # calculate Faith's Eveness as PD
sample_data(hiv_pos)$PD <- phylo_dist_pos$PD # add PD to metadata table
pd_pos <- sample_data(hiv_pos)$PD
hiv_pos_fr$PD <- pd_pos
wilcox_pos_labor <- wilcox.test(PD ~ Labor_category, data=hiv_pos_fr, exact=FALSE)$p.value
wilcox_pos_labor #p=0.4231
plot.pd_pos <- ggplot(sample_data(hiv_pos), aes(Labor_category, PD,fill=Labor_category)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("Labor vs non-labor of HIV-")+
  xlab("Labor category") +
  ylab("Eveness") +
  geom_signif(comparisons = list(c("Labor","Non-Labor")),
              map_signif_level= TRUE, test ="wilcox.test", y_position = 38)+ # plot any metadata category against the PD
  theme_minimal()
plot.pd_pos # view plot
ggsave(filename = "plot_PD_hivpos.png"
       , plot.pd_pos
       , height=4, width=6)

```

### HIV  status within labor
```{r}

phylo_dist_labor <- pd(t(otu_table(hiv_labor)), phy_tree(hiv_labor),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_labor)$PD <- phylo_dist_labor$PD # add PD to metadata table
pd_labor <- sample_data(hiv_labor)$PD
hiv_labor_fr$PD <- pd_labor
wilcox.labor <- wilcox.test(PD ~ HIV_Status, data=hiv_labor_fr, exact=FALSE)$p.value
wilcox.labor #p=0.1141
plot.pd_labor <- ggplot(sample_data(hiv_labor), aes(HIV_Status, PD,fill=HIV_Status)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("HIV+ vs HIV- of labor")+
  xlab("HIV Status") +
  ylab("phylogenetic diversity") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              map_signif_level= TRUE, test ="wilcox.test", y_Negition = 35)+ # plot any metadata category against the PD
  theme_minimal()
plot.pd_labor
ggsave(filename = "plot_PD_labor.png"
       , plot.pd_labor
       , height=4, width=6)

```

### HIV status within non labor
```{r}
phylo_dist_nonlabor <- pd(t(otu_table(hiv_nonlabor)), phy_tree(hiv_nonlabor),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_nonlabor)$PD <- phylo_dist_nonlabor$PD # add PD to metadata table
pd_nonlabor <- sample_data(hiv_nonlabor)$PD
hiv_nonlabor_fr$PD <- pd_nonlabor #add PD to dataframe
wilcox_nonlabor <- wilcox.test(PD ~ HIV_Status, data=hiv_nonlabor_fr, exact=FALSE)$p.value
wilcox_nonlabor #p=0.025
plot.pd_nonlabor <- ggplot(sample_data(hiv_nonlabor), aes(HIV_Status, PD,fill=HIV_Status)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("HIV+ vs HIV- of nonlabor")+
  xlab("HIV Status") +
  ylab("PD") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              map_signif_level= TRUE, test ="wilcox.test", y_Negition = 40)+ # plot any metadata category against the PD
  theme_minimal()
plot.pd_nonlabor
ggsave(filename = "plot_PD_nonlabor.png"
       , plot.pd_nonlabor
       , height=4, width=6)
```
### FDR
```{r}
p.values <- c(wilcox_nonlabor, wilcox.labor) #p=0.025, p=0.1141
p_adj <- p.adjust(p.values, method ="fdr")
p_adj
```
## Combined
```{r}
#combine data frames
combined <- rbind(hiv_pos_fr, hiv_pos_fr, hiv_labor_fr, hiv_nonlabor_fr)
combined_plot <- ggplot(combined, aes(x=interaction(HIV_Status,Labor_category),y= PD, fill =Labor_category)) +
  geom_violin(trim=FALSE, alpha =0.6,position=position_dodge(0.9))+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA, position=position_dodge(0.9)) +
  geom_jitter(shape=16, position=position_jitterdodge(jitter.width=0.2, dodge.width=0.9), size=1.5, alpha=0.6)+
  scale_fill_manual(values=c("#d7191c", "#fdae61", "#ffffbf","#abd9e9"))+
  labs(x="HIV Status",
       y= "PD",
       title="Faith's PD in HIV status and Labor category of Zimbabwe's patients") +
   geom_signif(comparisons = list(c("Negative.Labor","Positive.Labor"),c("Negative.Nonlabor","Positive.Nonlabor"),c("Positive.Labor","Positive.Nonlabor"),c("Negative.Labor","Negative.Nonlabor")), map_signif_level= TRUE, test ="wilcox.test") +
  theme_minimal()
combined_plot
ggsave(filename = "plot_PD_labor_status.png"
       , combined_plot
       , height=4, width=6)
```

## Education level

### Primary

```{r}
phylo_dist_primary <- pd(t(otu_table(hiv_primary)), phy_tree(hiv_primary),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_primary)$PD <- phylo_dist_primary$PD # add PD to metadata table
pd_primary <- sample_data(hiv_primary)$PD
hiv_primary_fr$PD <- pd_primary
wilcox_primary <- wilcox.test(PD ~ HIV_Status, data=hiv_primary_fr, exact=FALSE)$p.value #p=0.9584
plot.pd_primary <- ggplot(sample_data(hiv_primary), aes(x=HIV_Status, y=PD,fill= HIV_Status)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("HIV+ vs HIV- of Primary")+
  xlab("HIV Status") +
  ylab("PD") +
 geom_signif(comparisons = list(c("Negative","Positive")),
              map_signif_level= TRUE, test ="wilcox.test", y_position = 40)+
  theme_minimal()
plot.pd_primary
ggsave(filename = "plot_PD_primary.png"
       , plot.pd_primary
       , height=4, width=6)
```

### Secondary 
```{r}
phylo_dist_secondary <- pd(t(otu_table(hiv_secondary)), phy_tree(hiv_secondary),
                 include.root=F) # calculate Faith's Phylogenetic Diversity as PD
sample_data(hiv_secondary)$PD <- phylo_dist_secondary$PD # add PD to metadata table
pd_secondary <- sample_data(hiv_secondary)$PD
hiv_secondary_fr$PD <- pd_secondary
wilcox_secondary <- wilcox.test(PD ~ HIV_Status, data=hiv_secondary_fr, exact=FALSE)$p.value
wilcox_secondary#p=0.0025
plot.pd_secondary <- ggplot(sample_data(hiv_secondary), aes(x=HIV_Status, y=PD,fill= HIV_Status)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("HIV+ vs HIV- of secondary")+
  xlab("HIV Status") +
  ylab("PD") +
 geom_signif(comparisons = list(c("Negative","Positive")),
              map_signif_level= TRUE, test ="wilcox.test", y_position = 40)+
  theme_minimal()
plot.pd_secondary
ggsave(filename = "plot_PD_secondary.png"
       , plot.pd_secondary
       , height=4, width=6)
```
### Tertiary
```{r}
phylo_dist_tertiary <- pd(t(otu_table(hiv_tertiary)), phy_tree(hiv_tertiary),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_tertiary)$PD <- phylo_dist_tertiary$PD # add PD to metadata table
pd_tertiary <- sample_data(hiv_tertiary)$PD
hiv_tertiary_fr$PD <- pd_tertiary
wilcox_tertiary <- wilcox.test(PD ~ HIV_Status, data=hiv_tertiary_fr, exact=FALSE)
wilcox_tertiary #p=0.8597
plot.pd_tertiary <- ggplot(sample_data(hiv_tertiary), aes(HIV_Status, PD,fill= HIV_Status)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("HIV+ vs HIV- of tertiary")+
  xlab("HIV Status") +
  ylab("phylogenetic diversity") +
 geom_signif(comparisons = list(c("Negative","Positive")),
              map_signif_level= TRUE, test ="wilcox.test", y_position = 40)+
  theme_minimal()
plot.pd_tertiary
ggsave(filename = "plot_PD_tertiary.png"
       , plot.pd_tertiary
       , height=4, width=6)
```
## Age
### Young
```{r}
phylo_dist_young <- pd(t(otu_table(hiv_young)), phy_tree(hiv_young),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_young)$PD <- phylo_dist_young$PD # add PD to metadata table
pd_young <- sample_data(hiv_young)$PD
hiv_young_fr$PD <- pd_young
wilcox_young <- wilcox.test(PD ~ HIV_Status, data=hiv_young_fr, exact=FALSE)$p.value
wilcox_young #p=0.0078
plot.pd_young <- ggplot(sample_data(hiv_young), aes(HIV_Status, PD,fill= HIV_Status)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("HIV+ vs HIV- of young")+
  xlab("HIV Status") +
  ylab("phylogenetic diversity") +
 geom_signif(comparisons = list(c("Negative","Positive")),
              map_signif_level= TRUE, test ="wilcox.test", y_position = 35)+
  theme_minimal()
plot.pd_young
ggsave(filename = "plot_PD_young.png"
       , plot.pd_young
       , height=4, width=6)
```
### Old

```{r}
phylo_dist_old <- pd(t(otu_table(hiv_old)), phy_tree(hiv_old),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_old)$PD <- phylo_dist_old$PD # add PD to metadata table
pd_old <- sample_data(hiv_old)$PD
hiv_old_fr$PD <- pd_old
wilcox_old <- wilcox.test(PD ~ HIV_Status, data=hiv_old_fr, exact=FALSE)
wilcox_old #p=0.7953
plot.pd_old <- ggplot(sample_data(hiv_old), aes(HIV_Status, PD,fill= HIV_Status)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("HIV+ vs HIV- of old")+
  xlab("HIV Status") +
  ylab("phylogenetic diversity") +
 geom_signif(comparisons = list(c("Negative","Positive")),
              map_signif_level= TRUE, test ="wilcox.test", y_position = 40)+
  theme_minimal()
plot.pd_old
ggsave(filename = "plot_PD_old.png"
       , plot.pd_old
       , height=4, width=6)
```

### education within HIV+
```{r}
lm_ed <- lm(log(PD) ~ Education_Level,data = hiv_pos_fr) #p=0.4078
anova_ed <- aov(lm_ed)
summary(anova_ed)
TukeyHSD(anova_ed)
plot.pd_pos_ed <- ggplot(sample_data(hiv_pos), aes(x=Education_Level, y=PD,fill=Education_Level)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61","#ffffbf"))+
  ggtitle("Primary, Secondary, Tertiary of HIV+")+
  xlab("Education Level") +
  ylab("PD") +
  geom_signif(comparisons = list(c("Primary","Secondary"),c("Secondary","Tertiary"),c("Primary","Tertiary")),
              y_position = c(35, 37, 40),
annotations = c("NS.","NS.","NS."))+
  theme_minimal()# plot any metadata category against the PD
plot.pd_pos_ed # view plot
ggsave(filename = "plot_PD_hivpos_ed.png"
       , plot.pd_pos_ed
       , height=4, width=6)
```
### ed within HIV Negative 
```{r}
lm_ed_neg <- lm(log(PD) ~ Education_Level,data = hiv_neg_fr) #p=0.4078
anova_ed_neg <- aov(lm_ed_neg)
summary(anova_ed_neg)
TukeyHSD(anova_ed_neg)
plot.pd_neg_ed <- ggplot(sample_data(hiv_neg), aes(x=Education_Level, y=PD,fill=Education_Level)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61","#ffffbf"))+
  ggtitle("Primary, Secondary, Tertiary of HIV-")+
  xlab("Education Level") +
  ylab("PD") +
  geom_signif(comparisons = list(c("Primary","Secondary"),c("Secondary","Tertiary"),c("Primary","Tertiary")),
              y_negition = c(41, 45, 50), annotations = c("NS.","NS.","NS.")) +
  theme_minimal()# plot any metadata category against the PD
plot.pd_neg_ed # view plot
ggsave(filename = "plot_PD_hivneg_ed.png"
       , plot.pd_neg_ed
       , height=4, width=6)
```

## Age within HIV-
```{r}
wilcox_neg_age <- wilcox.test(PD ~ Visit_Age, data=hiv_neg_fr, exact=FALSE)
wilcox_neg_age #p=0.2387
plot.pd_neg_age <- ggplot(sample_data(hiv_neg), aes(Visit_Age, PD,fill=Visit_Age)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("<45, >45yo of HIV-")+
  geom_signif(comparisons = list(c("Younger than 45","Older than 45")),
              map_signif_level= TRUE, test ="wilcox.test", y_negition = 40)+
  theme_minimal() # plot any metadata category against the PD
plot.pd_neg_age # view plot
ggsave(filename = "plot_PD_hivneg_age.png"
       , plot.pd_neg_age
       , height=4, width=6)

```

### Age within HIV+
```{r}
wilcox_pos_age <- wilcox.test(PD ~ Visit_Age, data=hiv_pos_fr, exact=FALSE)
wilcox_pos_age #p=0.09295
plot.pd_pos_age <- ggplot(sample_data(hiv_pos), aes(Visit_Age, PD,fill=Visit_Age)) + 
  geom_violin(trim=FALSE, alpha =0.6)+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA) +
  geom_jitter(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  ggtitle("<45, >45yo of HIV+")+
  geom_signif(comparisons = list(c("Younger than 45","Older than 45")),
              map_signif_level= TRUE, test ="wilcox.test", y_position = 40)+
  theme_minimal()
plot.pd_pos_age # view plot
ggsave(filename = "plot_PD_hivpos_age.png"
       , plot.pd_pos_age
       , height=4, width=6)

```

## Age Combined
```{r}
#combine data frames, Positive old and young signifi
combined_age <- rbind(hiv_pos_fr, hiv_pos_fr, hiv_young_fr, hiv_old_fr)
combined_plot_age <- ggplot(combined, aes(x=interaction(HIV_Status,Visit_Age),y= PD, fill =Visit_Age)) +
  geom_violin(trim=FALSE, alpha =0.6,position=position_dodge(0.9))+
  geom_boxplot(width=0.1, fill='white', outlier.shape=NA, position=position_dodge(0.9)) +
  geom_jitter(shape=16, position=position_jitterdodge(jitter.width=0.2, dodge.width=0.9), size=1.5, alpha=0.6)+
  scale_fill_manual(values=c("#d7191c", "#fdae61", "#ffffbf","#abd9e9"))+
  labs(x="HIV Status",
       y= "Faith's Eveness",
       title="Faith's PD in HIV status and Age of Zimbabwe's patients") +
  theme_minimal()
combined_plot_age
ggsave(filename = "plot_PD_age_combined.png"
       , combined_plot_age
       , height=4, width=6)
```
### Education combined
```{r}
#combine data frames, Positive old and young signifi
combined_ed <- rbind(hiv_pos_fr, hiv_pos_fr, hiv_primary_fr, hiv_secondary_fr, hiv_tertiary_fr)
combined_plot_ed <- ggplot(combined_ed, aes(x=HIV_Status, y=PD, fill =Education_Level)) +
  geom_boxplot() +
  scale_fill_manual(values= c("#d7191c","#fdae61","#ffffbf")) +
  labs(x="HIV Status",
       y= "Faith's Eveness",
       title="Faith's PD in HIV status and ed of Zimbabwe's patients") +
  theme_bw()
combined_plot_ed
ggsave(filename = "plot_PD_ed_combined.png"
       , combined_plot_ed
       , height=4, width=6)
```


#Beta diversity #

## HIV postive: labor cat
```{r}

bc_dm_pos <- distance(hiv_pos, method="wunifrac")
adonis2(bc_dm_pos ~ Labor_category, data=hiv_pos_fr)
# check which methods you can specify
pcoa_bc_pos <- ordinate(hiv_pos, method="PCoA", distance=bc_dm_pos)
gg_pcoa_pos <- plot_ordination(hiv_pos, pcoa_bc_pos, color = "Labor_category") 
pcoa_pos = gg_pcoa_pos + #change this
  geom_point(size = 3)+ 
  labs(title ="Beta diversity within labor categories of HIV Positive patients") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_pos

ggsave(filename = "wunifrac_hivpos_labor.png"
       , pcoa_pos
       , height=4, width=6)
```

## hiv negative: labor cat ##
```{r}
bc_dm_neg <- distance(hiv_neg, method="wunifrac")
# check which methods you can specify
pcoa_bc_neg <- ordinate(hiv_neg, method="PCoA", distance=bc_dm_neg)
gg_pcoa_neg <- plot_ordination(hiv_neg, pcoa_bc_neg, color = "Labor_category") 
pcoa_neg = gg_pcoa_neg + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity within labor categories of HIV negative patients") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_neg

ggsave(filename = "wunifrac_hivneg_labor.png"
       , pcoa_neg
       , height=4, width=6)
```

## labor : hiv status ##
```{r}
bc_dm_labor <- distance(hiv_labor, method="wunifrac")
# check which methods you can specify
pcoa_bc_labor <- ordinate(hiv_labor, method="PCoA", distance=bc_dm_labor)
gg_pcoa_labor <- plot_ordination(hiv_labor, pcoa_bc_labor, color = "HIV_Status") 
pcoa_labor = gg_pcoa_labor + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ labor and HIV- labor") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_labor
ggsave(filename = "wunifrac_hiv_labor.png"
       , pcoa_labor
       , height=4, width=6)
```


## nonlabor ##
```{r}
bc_dm_nonlabor <- distance(hiv_nonlabor, method="wunifrac")
# check which methods you can specify
pcoa_bc_nonlabor <- ordinate(hiv_nonlabor, method="PCoA", distance=bc_dm_nonlabor)
gg_pcoa_nonlabor <- plot_ordination(hiv_nonlabor, pcoa_bc_nonlabor, color = "HIV_Status") 
pcoa_nonlabor = gg_pcoa_nonlabor + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ nonlabor and HIV- nonlabor") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_nonlabor

ggsave(filename = "wunifrac_hiv_nonlabor.png"
       , pcoa_nonlabor
       , height=4, width=6)
```
## HIV+ : Age visit
```{r}
adonis2(bc_dm_pos ~ Visit_Age, data=hiv_pos_fr)
gg_pcoa_pos_age <- plot_ordination(hiv_pos, pcoa_bc_pos, color = "Visit_Age") 
pcoa_pos_age = gg_pcoa_pos_age + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ old and HIV+ young") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_pos_age

ggsave(filename = "wunifrac_hiv_pos_age.png"
       , pcoa_pos_age
       , height=4, width=6)
```
## HIV-: Age visit
```{r}
gg_pcoa_neg_age <- plot_ordination(hiv_neg, pcoa_bc_neg, color = "Visit_Age") 
pcoa_neg_age = gg_pcoa_neg_age + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV- old and HIV- young") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_neg_age

ggsave(filename = "wunifrac_hiv_neg_age.png"
       , pcoa_neg_age
       , height=4, width=6)
```
## HIV+ : Education
```{r}
adonis2(bc_dm_pos ~ Education_Level, data=hiv_pos_fr)
pcoa_bc_pos <- ordinate(hiv_pos, method = "PCoA", distance = bc_dm_pos)
gg_pcoa_pos_ed <- plot_ordination(hiv_pos, pcoa_bc_pos, color = "Education_Level") 
# Get unique pairwise group combinations
edu_levels <- unique(hiv_pos_fr$Education_Level)
pairwise_combos <- combn(edu_levels, 2, simplify = FALSE)

# Run adonis2 for each pair
pairwise_results <- lapply(pairwise_combos, function(pair) {
  # Subset metadata
  keep <- hiv_pos_fr$Education_Level %in% pair
  sub_meta <- hiv_pos_fr[keep, ]
  
  # Subset distance matrix
  sub_dm <- as.matrix(bc_dm_pos)[keep, keep]
  
  # Run PERMANOVA
  res <- adonis2(as.dist(sub_dm) ~ Education_Level, data = sub_meta)
  
  data.frame(
    Group1 = pair[1],
    Group2 = pair[2],
    R2 = res$R2[1],
    F = res$F[1],
    p = res$`Pr(>F)`[1]
  )
})

# Combine into a table
do.call(rbind, pairwise_results)

pcoa_pos_ed = gg_pcoa_pos_ed + 
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ primary, secondary, tertiary") +
  scale_color_manual(values=c("#d7191c","#fdae61","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "t")
pcoa_pos_ed

ggsave(filename = "wunifrac_hiv_pos_ed.png"
       , pcoa_pos_ed
       , height=4, width=6)
```
## HIV- : Education
```{r}
gg_pcoa_neg_ed <- plot_ordination(hiv_neg, pcoa_bc_neg, color = "Education_Level") 
pcoa_neg_ed = gg_pcoa_neg_ed + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV- primary, secondary, tertiary") +
  scale_color_manual(values=c("#d7191c","#fdae61","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_neg_ed

ggsave(filename = "wunifrac_hiv_neg_ed.png"
       , pcoa_neg_ed
       , height=4, width=6)
```
## Young ##
```{r}
bc_dm_young <- distance(hiv_young, method="wunifrac")
# check which methods you can specify
pcoa_bc_young <- ordinate(hiv_young, method="PCoA", distance=bc_dm_young)
gg_pcoa_young <- plot_ordination(hiv_young, pcoa_bc_young, color = "HIV_Status") 
pcoa_young = gg_pcoa_young + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ young and HIV- young") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_young

ggsave(filename = "wunifrac_hiv_young.png"
       , pcoa_young
       , height=4, width=6)
```
## Old ##
```{r}
bc_dm_old <- distance(hiv_old, method="wunifrac")
# check which methods you can specify
pcoa_bc_old <- ordinate(hiv_old, method="PCoA", distance=bc_dm_old)
gg_pcoa_old <- plot_ordination(hiv_old, pcoa_bc_old, color = "HIV_Status") 
pcoa_old = gg_pcoa_old + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ old and HIV- old") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
   theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_old

ggsave(filename = "wunifrac_hiv_old.png"
       , pcoa_old
       , height=4, width=6)
```

## Primary ##
```{r}
bc_dm_primary <- distance(hiv_primary, method="wunifrac")
# check which methods you can specify
pcoa_bc_primary <- ordinate(hiv_primary, method="PCoA", distance=bc_dm_primary)
gg_pcoa_primary <- plot_ordination(hiv_primary, pcoa_bc_primary, color = "HIV_Status") 
pcoa_primary = gg_pcoa_primary + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ primary and HIV- primary") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_primary

ggsave(filename = "wunifrac_hiv_primary.png"
       , pcoa_primary
       , height=4, width=6)
```

## Secondary ##
```{r}
bc_dm_secondary <- distance(hiv_secondary, method="wunifrac")
# check which methods you can specify
pcoa_bc_secondary <- ordinate(hiv_secondary, method="PCoA", distance=bc_dm_secondary)
gg_pcoa_secondary <- plot_ordination(hiv_secondary, pcoa_bc_secondary, color = "HIV_Status") 
pcoa_secondary = gg_pcoa_secondary + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ secondary and HIV- secondary") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
   theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_secondary
ggsave(filename = "wunifrac_hiv_secondary.png"
       , pcoa_secondary
       , height=4, width=6)
```
## tertiary ##
```{r}
bc_dm_tertiary <- distance(hiv_tertiary, method="wunifrac")
# check which methods you can specify
pcoa_bc_tertiary <- ordinate(hiv_tertiary, method="PCoA", distance=bc_dm_tertiary)
gg_pcoa_tertiary <- plot_ordination(hiv_tertiary, pcoa_bc_tertiary, color = "HIV_Status") 
pcoa_tertiary = gg_pcoa_tertiary + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ tertiary and HIV- tertiary") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
   theme_minimal()+
  stat_ellipse(type = "norm")
pcoa_tertiary
ggsave(filename = "wunifrac_hiv_tertiary.png"
       , pcoa_tertiary
       , height=4, width=6)
```
