---
title: "Alpha-beta-labortype"
author: "Tien Do"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r include=FALSE}
library(phyloseq)
library(ape)
library(tidyverse)
library(picante)
library(ggsignif)
library(ggplot2)
library("car")
```
#### Load in RData ####
```{r}
load("hiv_rare.RData")
load("hiv_final.RData")
```

#### Filter metadata ####
```{r}
#duplicate phylosesq object
phyloseq_hiv <- hiv_rare 
# Filter out 3rd visit samples:
phyloseq_filter_hiv <- subset_samples(phyloseq_hiv,Visit_Cat == "2nd Visit")
# change labor categories
labor_rename <- as.data.frame(sample_data(phyloseq_filter_hiv))
```
```{r}
labor_rename$Visit_Age <- ifelse(labor_rename$Visit_Age > 45, "Older than 45", "Younger than 45")
labor_rename$Labor_category <- ifelse(labor_rename$Labor_category %in% c("Non_Labor_Houseworker","Non_Labor_Other","Unemployed"), "Non-Labor", "Labor")
sample_data(phyloseq_filter_hiv) <- sample_data(labor_rename)
```
```{r}
#extract the sample data:
sample_phyloseq_hiv <- sample_data(phyloseq_filter_hiv)
#retain wanted columns:
phyloseq_labor<- sample_phyloseq_hiv[, c( "HIV_Status", "Labor_category","Education_Level","Visit_Age")]
# Replace the sample data in your phyloseq object with the updated one
sample_data(phyloseq_filter_hiv) <- phyloseq_labor
#filter out 
hiv_pos <- subset_samples(phyloseq_filter_hiv, HIV_Status == "Positive")
hiv_pos_fr <- sample_data(hiv_pos)
hiv_pos_fr <- data.frame(hiv_pos_fr)
hiv_neg <- subset_samples(phyloseq_filter_hiv, HIV_Status =="Negative")
hiv_neg_fr <- sample_data(hiv_neg)
hiv_neg_fr <- data.frame(hiv_neg_fr)
hiv_labor <- subset_samples(phyloseq_filter_hiv, Labor_category == "Labor")
hiv_labor_fr <- sample_data(hiv_labor)
hiv_labor_fr <- data.frame(hiv_labor_fr)
hiv_nonlabor <- subset_samples(phyloseq_filter_hiv, Labor_category == "Non-Labor")
hiv_nonlabor_fr <- sample_data(hiv_nonlabor)
hiv_nonlabor_fr <- data.frame(hiv_nonlabor_fr)
#filter education level: tertiary, secondary, tertiary
hiv_primary <- subset_samples(phyloseq_filter_hiv, Education_Level == "Primary")
hiv_primary_fr <- sample_data(hiv_primary)
hiv_primary_fr <- data.frame(hiv_primary_fr)
hiv_secondary <- subset_samples(phyloseq_filter_hiv, Education_Level == "Secondary")
hiv_secondary_fr <- sample_data(hiv_secondary)
hiv_secondary_fr <- data.frame(hiv_secondary_fr)
hiv_tertiary <- subset_samples(phyloseq_filter_hiv, Education_Level == "Tertiary")
hiv_tertiary_fr <- sample_data(hiv_tertiary)
hiv_tertiary_fr <- data.frame(hiv_tertiary_fr)
hiv_young <- subset_samples(phyloseq_filter_hiv, Visit_Age == "Younger than 45")
hiv_young_fr <- sample_data(hiv_young)
hiv_young_fr <- data.frame(hiv_young_fr)
hiv_old <- subset_samples(phyloseq_filter_hiv, Visit_Age == "Older than 45")
hiv_old_fr <- sample_data(hiv_old)
hiv_old_fr <- data.frame(hiv_old_fr)
```


# Alpha diversity #


## phylogenetic diversity ##
### Labor type within HIV+
```{r}

phylo_dist_pos <- pd(t(otu_table(hiv_pos)), phy_tree(hiv_pos),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_pos)$PD <- phylo_dist_pos$PD # add PD to metadata table
pd_pos <- sample_data(hiv_pos)$PD
hiv_pos_fr$PD <- pd_pos
wilcox_pos_labor <- wilcox.test(PD ~ Labor_category, data=hiv_pos_fr, exact=FALSE) #p=0.4078
plot.pd_pos <- ggplot(sample_data(hiv_pos), aes(Labor_category, PD,fill=HIV_Status)) + 
  geom_boxplot() +
  geom_point(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  xlab("Labor category") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Labor","Non-Labor")),
              annotations = "ns") # plot any metadata category against the PD

plot.pd_pos # view plot
ggsave(filename = "plot_PD_hivpos.png"
       , plot.pd_pos
       , height=4, width=6)
```
### Labor type within HIV-
```{r}
phylo_dist_neg <- pd(t(otu_table(hiv_neg)), phy_tree(hiv_neg),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_neg)$PD <- phylo_dist_neg$PD # add PD to metadata table
pd_neg <- sample_data(hiv_neg)$PD
hiv_neg_fr$PD <- pd_neg
wilcox_neg_labor <- wilcox.test(PD ~ Labor_category, data=hiv_neg_fr, exact=FALSE)
wilcox_neg_labor #p=0.4231
plot.pd_neg <- ggplot(sample_data(hiv_neg), aes(Labor_category, PD,fill=HIV_Status)) + 
  geom_boxplot() +
  geom_point(size=2)+
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  xlab("Labor category") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Labor","Non-Labor")),
              annotations = "ns") # plot any metadata category against the PD
plot.pd_neg # view plot
ggsave(filename = "plot_PD_hivneg.png"
       , plot.pd_neg
       , height=4, width=6)

```

### HIV  status within labor
```{r}

phylo_dist <- pd(t(otu_table(hiv_labor)), phy_tree(hiv_labor),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_labor)$PD <- phylo_dist$PD # add PD to metadata table
pd_labor <- sample_data(hiv_labor)$PD
hiv_labor_fr$PD <- pd_labor
wilcox_labor_labor <- wilcox.test(PD ~ HIV_Status, data=hiv_labor_fr, exact=FALSE)
wilcox_labor_labor #p=0.1141
plot.pd_labor <- ggplot(sample_data(hiv_labor), aes(HIV_Status, PD,fill=Labor_category)) + 
  geom_boxplot() +
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  geom_point(size=2)+
  xlab("HIV Status") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              annotations = "ns") # plot any metadata category against the PD
ggsave(filename = "plot_PD_labor.png"
       , plot.pd_labor
       , height=4, width=6)

```

### HIV status within non labor
```{r}
phylo_dist <- pd(t(otu_table(hiv_nonlabor)), phy_tree(hiv_nonlabor),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_nonlabor)$PD <- phylo_dist$PD # add PD to metadata table
pd_nonlabor <- sample_data(hiv_nonlabor)$PD
hiv_nonlabor_fr$PD <- pd_nonlabor #add PD to dataframe
wilcox_nonlabor <- wilcox.test(PD ~ HIV_Status, data=hiv_nonlabor_fr, exact=FALSE)
wilcox_nonlabor #p=0.1141
plot.pd_nonlabor <- ggplot(sample_data(hiv_nonlabor), aes(HIV_Status, PD, fill=Labor_category)) + 
  geom_boxplot() +
  scale_fill_manual(values= c("#d7191c", "#fdae61"))+
  geom_point(size=2)+
  xlab("HIV Status") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              annotations = "0.025") # plot any metadata category against the PD
plot.pd_nonlabor # view plot
ggsave(filename = "plot_PD_nonlabor.png"
       , plot.pd_nonlabor
       , height=4, width=6)
```
## Combined
```{r}
#combine data frames
combined <- rbind(hiv_pos_fr, hiv_neg_fr, hiv_labor_fr, hiv_nonlabor_fr)
combined_plot <- ggplot(combined, aes(HIV_Status, PD, fill =Labor_category)) +
  geom_boxplot() +
  scale_fill_manual(values=c("#d7191c", "#fdae61"))+
  labs(x="HIV Status",
       y= "Faith's Phylogenetic Diversity",
       title="Faith's PD in HIV status and Labor category of Zimbabwe's patients") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              annotations = "0.025")
ggsave(filename = "plot_PD_labor_status.png"
       , combined_plot
       , height=4, width=6)
```

## Education level

### Primary

```{r}
phylo_dist_primary <- pd(t(otu_table(hiv_primary)), phy_tree(hiv_primary),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_primary)$PD <- phylo_dist_primary$PD # add PD to metadata table
pd_primary <- sample_data(hiv_primary)$PD
hiv_primary_fr$PD <- pd_primary
wilcox_primary <- wilcox.test(PD ~ HIV_Status, data=hiv_primary_fr, exact=FALSE)
wilcox_primary #p=0.9584
plot.pd_primary <- ggplot(sample_data(hiv_primary), aes(HIV_Status, PD,fill= Education_Level)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("HIV Status") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              annotations = "ns") # plot any metadata category against the PD
ggsave(filename = "plot_PD_primary.png"
       , plot.pd_primary
       , height=4, width=6)
```

### Secondary 
```{r}
phylo_dist_secondary <- pd(t(otu_table(hiv_secondary)), phy_tree(hiv_secondary),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_secondary)$PD <- phylo_dist_secondary$PD # add PD to metadata table
pd_secondary <- sample_data(hiv_secondary)$PD
hiv_secondary_fr$PD <- pd_secondary
wilcox_secondary <- wilcox.test(PD ~ HIV_Status, data=hiv_secondary_fr, exact=FALSE)$p.value
wilcox_secondary#p=0.0025
p = p.adjust(wilcox_secondary, method = "fdr") #padj=0.0025
plot.pd_secondary <- ggplot(sample_data(hiv_secondary), aes(HIV_Status, PD,fill= Education_Level)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("HIV Status") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              annotations = "0.0025") # plot any metadata category against the PD
plot.pd_secondary
ggsave(filename = "plot_PD_secondary.png"
       , plot.pd_secondary
       , height=4, width=6)
```
### Tertiary
```{r}
phylo_dist_tertiary <- pd(t(otu_table(hiv_tertiary)), phy_tree(hiv_tertiary),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_tertiary)$PD <- phylo_dist_tertiary$PD # add PD to metadata table
pd_tertiary <- sample_data(hiv_tertiary)$PD
hiv_tertiary_fr$PD <- pd_tertiary
wilcox_tertiary <- wilcox.test(PD ~ HIV_Status, data=hiv_tertiary_fr, exact=FALSE)
wilcox_tertiary #p=0.8597
plot.pd_tertiary <- ggplot(sample_data(hiv_tertiary), aes(HIV_Status, PD,fill= Education_Level)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("HIV Status") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              annotations = "ns") # plot any metadata category against the PD
plot.pd_tertiary
ggsave(filename = "plot_PD_tertiary.png"
       , plot.pd_tertiary
       , height=4, width=6)
```
## Age
### Young
```{r}
phylo_dist_young <- pd(t(otu_table(hiv_young)), phy_tree(hiv_young),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_young)$PD <- phylo_dist_young$PD # add PD to metadata table
pd_young <- sample_data(hiv_young)$PD
hiv_young_fr$PD <- pd_young
wilcox_young <- wilcox.test(PD ~ HIV_Status, data=hiv_young_fr, exact=FALSE)$p.value
wilcox_young #p=0.0078
p.adjust(wilcox_young, "fdr")
plot.pd_young <- ggplot(sample_data(hiv_young), aes(HIV_Status, PD,fill= Visit_Age)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("HIV Status") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              annotations = "0.0078") # plot any metadata category against the PD
plot.pd_young
ggsave(filename = "plot_PD_young.png"
       , plot.pd_young
       , height=4, width=6)
```
### Old

```{r}
phylo_dist_old <- pd(t(otu_table(hiv_old)), phy_tree(hiv_old),
                 include.root=F) # calculate Faith's phylogenetic diversity as PD
sample_data(hiv_old)$PD <- phylo_dist_old$PD # add PD to metadata table
pd_old <- sample_data(hiv_old)$PD
hiv_old_fr$PD <- pd_old
wilcox_old <- wilcox.test(PD ~ HIV_Status, data=hiv_old_fr, exact=FALSE)
wilcox_old #p=0.7953
plot.pd_old <- ggplot(sample_data(hiv_old), aes(HIV_Status, PD,fill= Visit_Age)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("HIV Status") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Negative","Positive")),
              annotations = "ns") # plot any metadata category against the PD
plot.pd_old
ggsave(filename = "plot_PD_old.png"
       , plot.pd_old
       , height=4, width=6)
```

### education within HIV+
```{r}
lm_ed <- lm(log(PD) ~ Education_Level,data = hiv_pos_fr) #p=0.4078
anova_ed <- aov(lm_ed)
summary(anova_ed)
TukeyHSD(anova_ed)
plot.pd_pos_ed <- ggplot(sample_data(hiv_pos), aes(Education_Level, PD,fill=HIV_Status)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("Education Level") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Primary","Secondary"),c("Secondary","Tertiary"),c("Primary","Tertiary")),
              y_position = c(28, 28, 30),
annotations = c("ns","ns","ns")) # plot any metadata category against the PD
plot.pd_pos_ed
plot.pd_pos_ed # view plot
ggsave(filename = "plot_PD_hivpos_ed.png"
       , plot.pd_pos_ed
       , height=4, width=6)
```
### ed within HIV negative 
```{r}
lm_ed_neg <- lm(log(PD) ~ Education_Level,data = hiv_neg_fr) #p=0.4078
anova_ed_neg <- aov(lm_ed_neg)
summary(anova_ed_neg)
TukeyHSD(anova_ed_neg)
plot.pd_neg_ed <- ggplot(sample_data(hiv_neg), aes(Education_Level, PD,fill=HIV_Status)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("Education Level") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Primary","Secondary"),c("Secondary","Tertiary"),c("Primary","Tertiary")),
              y_position = c(28, 28, 30),
annotations = c("ns","ns","ns")) # plot any metadata category against the PD
plot.pd_neg_ed # view plot
ggsave(filename = "plot_PD_hivneg_ed.png"
       , plot.pd_neg_ed
       , height=4, width=6)
```

## Age within HIV-
```{r}
wilcox_neg_age <- wilcox.test(PD ~ Visit_Age, data=hiv_neg_fr, exact=FALSE)
wilcox_neg_age #p=0.2387
plot.pd_neg_age <- ggplot(sample_data(hiv_neg), aes(Visit_Age, PD,fill=HIV_Status)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("Age Visit") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Young","Old")),
              annotations = "ns") # plot any metadata category against the PD
plot.pd_neg_age # view plot
ggsave(filename = "plot_PD_hivneg_age.png"
       , plot.pd_neg_age
       , height=4, width=6)

```

### Age within HIV+
```{r}
wilcox_pos_age <- wilcox.test(PD ~ Visit_Age, data=hiv_pos_fr, exact=FALSE)
wilcox_pos_age #p=0.09295
plot.pd_pos_age <- ggplot(sample_data(hiv_pos), aes(Visit_Age, PD,fill=HIV_Status)) + 
  geom_boxplot() +
  geom_point(size=2)+
  xlab("Age Visit") +
  ylab("Phylogenetic Diversity") +
  geom_signif(comparisons = list(c("Young","Old")),
              annotations = "0.09295") # plot any metadata category against the PD
plot.pd_pos_age # view plot
ggsave(filename = "plot_PD_hivpos_age.png"
       , plot.pd_pos_age
       , height=4, width=6)

```

## Age Combined
```{r}
#combine data frames, positive old and young signifi
combined_age <- rbind(hiv_pos_fr, hiv_neg_fr, hiv_young_fr, hiv_old_fr)
combined_plot_age <- ggplot(combined_age, aes(HIV_Status, PD, fill =Visit_Age)) +
  geom_boxplot() +
  labs(x="HIV Status",
       y= "Faith's Phylogenetic Diversity",
       title="Faith's PD in HIV status and Age of Zimbabwe's patients") 
combined_plot_age
ggsave(filename = "plot_PD_age_combined.png"
       , combined_plot_age
       , height=4, width=6)
```
### Education combined
```{r}
#combine data frames, positive old and young signifi
combined_ed <- rbind(hiv_pos_fr, hiv_neg_fr, hiv_primary_fr, hiv_secondary_fr, hiv_tertiary_fr)
combined_plot_ed <- ggplot(combined_ed, aes(x=HIV_Status, y=PD, fill =Education_Level)) +
  geom_boxplot() +
  scale_fill_manual(values= c("#d7191c","#fdae61","#ffffbf")) +
  labs(x="HIV Status",
       y= "Faith's Phylogenetic Diversity",
       title="Faith's PD in HIV status and ed of Zimbabwe's patients") +
  theme_bw()
combined_plot_ed
ggsave(filename = "plot_PD_ed_combined.png"
       , combined_plot_ed
       , height=4, width=6)
```


#### Beta diversity #####

## HIV postive: labor cat
```{r}

bc_dm_pos <- distance(hiv_pos, method="wunifrac")
# check which methods you can specify
pcoa_bc_pos <- ordinate(hiv_pos, method="PCoA", distance=bc_dm_pos)
gg_pcoa_pos <- plot_ordination(hiv_pos, pcoa_bc_pos, color = "Labor_category") 
pcoa_pos = gg_pcoa_pos + #change this
  geom_point(size = 3)+ 
  labs(title ="Beta diversity within labor categories of HIV positive patients") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_pos

ggsave(filename = "wunifrac_hivpos_labor.png"
       , pcoa_pos
       , height=4, width=6)
```

## hiv negative: labor cat ##
```{r}
bc_dm_neg <- distance(hiv_neg, method="wunifrac")
# check which methods you can specify
pcoa_bc_neg <- ordinate(hiv_neg, method="PCoA", distance=bc_dm_neg)
gg_pcoa_neg <- plot_ordination(hiv_neg, pcoa_bc_neg, color = "Labor_category") 
pcoa_neg = gg_pcoa_neg + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity within labor categories of HIV negative patients") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_neg

ggsave(filename = "wunifrac_hivneg_labor.png"
       , pcoa_neg
       , height=4, width=6)
```

## labor : hiv status ##
```{r}
bc_dm_labor <- distance(hiv_labor, method="wunifrac")
# check which methods you can specify
pcoa_bc_labor <- ordinate(hiv_labor, method="PCoA", distance=bc_dm_labor)
gg_pcoa_labor <- plot_ordination(hiv_labor, pcoa_bc_labor, color = "HIV_Status") 
pcoa_labor = gg_pcoa_labor + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ labor and HIV- labor") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_labor

ggsave(filename = "wunifrac_hiv_labor.png"
       , pcoa_labor
       , height=4, width=6)
```


## nonlabor ##
```{r}
bc_dm_nonlabor <- distance(hiv_nonlabor, method="wunifrac")
# check which methods you can specify
pcoa_bc_nonlabor <- ordinate(hiv_nonlabor, method="PCoA", distance=bc_dm_nonlabor)
gg_pcoa_nonlabor <- plot_ordination(hiv_nonlabor, pcoa_bc_nonlabor, color = "HIV_Status") 
pcoa_nonlabor = gg_pcoa_nonlabor + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ nonlabor and HIV- nonlabor") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_nonlabor

ggsave(filename = "wunifrac_hiv_nonlabor.png"
       , pcoa_nonlabor
       , height=4, width=6)
```
## HIV+ : Age visit
```{r}
gg_pcoa_pos_age <- plot_ordination(hiv_pos, pcoa_bc_pos, color = "Visit_Age") 
pcoa_pos_age = gg_pcoa_pos_age + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ old and HIV+ young") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_pos_age

ggsave(filename = "wunifrac_hiv_pos_age.png"
       , pcoa_pos_age
       , height=4, width=6)
```
## HIV-: Age visit
```{r}
gg_pcoa_neg_age <- plot_ordination(hiv_neg, pcoa_bc_neg, color = "Visit_Age") 
pcoa_neg_age = gg_pcoa_neg_age + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV- old and HIV- young") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_neg_age

ggsave(filename = "wunifrac_hiv_neg_age.png"
       , pcoa_neg_age
       , height=4, width=6)
```
## HIV+ : Education
```{r}
gg_pcoa_pos_ed <- plot_ordination(hiv_pos, pcoa_bc_pos, color = "Education_Level") 
pcoa_pos_ed = gg_pcoa_pos_ed + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ primary, secondary, tertiary") +
  scale_color_manual(values=c("#d7191c","#fdae61","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_pos_ed

ggsave(filename = "wunifrac_hiv_pos_ed.png"
       , pcoa_pos_ed
       , height=4, width=6)
```
## HIV- : Education
```{r}
gg_pcoa_neg_ed <- plot_ordination(hiv_neg, pcoa_bc_neg, color = "Education_Level") 
pcoa_neg_ed = gg_pcoa_neg_ed + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV- primary, secondary, tertiary") +
  scale_color_manual(values=c("#d7191c","#fdae61","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_neg_ed

ggsave(filename = "wunifrac_hiv_neg_ed.png"
       , pcoa_neg_ed
       , height=4, width=6)
```
## Young ##
```{r}
bc_dm_young <- distance(hiv_young, method="wunifrac")
# check which methods you can specify
pcoa_bc_young <- ordinate(hiv_young, method="PCoA", distance=bc_dm_young)
gg_pcoa_young <- plot_ordination(hiv_young, pcoa_bc_young, color = "HIV_Status") 
pcoa_young = gg_pcoa_young + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ young and HIV- young") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_young

ggsave(filename = "wunifrac_hiv_young.png"
       , pcoa_young
       , height=4, width=6)
```
## Old ##
```{r}
bc_dm_old <- distance(hiv_old, method="wunifrac")
# check which methods you can specify
pcoa_bc_old <- ordinate(hiv_old, method="PCoA", distance=bc_dm_old)
gg_pcoa_old <- plot_ordination(hiv_old, pcoa_bc_old, color = "HIV_Status") 
pcoa_old = gg_pcoa_old + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ old and HIV- old") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_old

ggsave(filename = "wunifrac_hiv_old.png"
       , pcoa_old
       , height=4, width=6)
```

## Primary ##
```{r}
bc_dm_primary <- distance(hiv_primary, method="wunifrac")
# check which methods you can specify
pcoa_bc_primary <- ordinate(hiv_primary, method="PCoA", distance=bc_dm_primary)
gg_pcoa_primary <- plot_ordination(hiv_primary, pcoa_bc_primary, color = "HIV_Status") 
pcoa_primary = gg_pcoa_primary + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ primary and HIV- primary") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_primary

ggsave(filename = "wunifrac_hiv_primary.png"
       , pcoa_primary
       , height=4, width=6)
```

## Secondary ##
```{r}
bc_dm_secondary <- distance(hiv_secondary, method="wunifrac")
# check which methods you can specify
pcoa_bc_secondary <- ordinate(hiv_secondary, method="PCoA", distance=bc_dm_secondary)
gg_pcoa_secondary <- plot_ordination(hiv_secondary, pcoa_bc_secondary, color = "HIV_Status") 
pcoa_secondary = gg_pcoa_secondary + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ secondary and HIV- secondary") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_secondary
ggsave(filename = "wunifrac_hiv_secondary.png"
       , pcoa_secondary
       , height=4, width=6)
```
## tertiary ##
```{r}
bc_dm_tertiary <- distance(hiv_tertiary, method="wunifrac")
# check which methods you can specify
pcoa_bc_tertiary <- ordinate(hiv_tertiary, method="PCoA", distance=bc_dm_tertiary)
gg_pcoa_tertiary <- plot_ordination(hiv_tertiary, pcoa_bc_tertiary, color = "HIV_Status") 
pcoa_tertiary = gg_pcoa_tertiary + #change this
  geom_point(size = 3)+ 
  labs(title = "Beta diversity between HIV+ tertiary and HIV- tertiary") +
  scale_color_manual(values=c("#d7191c","#2c7bb6"))+
  theme_bw()+
  stat_ellipse(type = "norm")
pcoa_tertiary

ggsave(filename = "wunifrac_hiv_tertiary.png"
       , pcoa_tertiary
       , height=4, width=6)
```
