---
title: "Alpha-beta-labor-type"
author: "Tien Do"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Convert data into phyloseq object, codes are taken from Canvas module
```{r echo = FALSE }
library(tidyverse)
library(phyloseq)
library(ape)
library(vegan)
library(ggplot2)
```

```{r} 
metafp <- "../QIIME2 analysis/hiv_metadata.tsv" #import metadata, otu table, taxonomy, and phylo tree
meta <- read_delim(metafp, delim="\t")
otufp <- "../QIIME2 analysis/table-hiv.txt"
otu <- read_delim(file = otufp, delim="\t", skip=1)

taxfp <- "../QIIME2 analysis/taxonomy.tsv"
tax <- read_delim(taxfp, delim="\t")

phylotreefp <- "../QIIME2 analysis/tree.nwk"
phylotree <- read.tree(phylotreefp)
```


#### Format OTU table ####
# OTU tables should be a matrix
# with rownames and colnames as OTUs and sampleIDs, respectively
# Note: tibbles do not allow rownames so if you imported with read_delim, change back
```{r}
# save everything except first column (OTU ID) into a matrix
otu_mat <- as.matrix(otu[,-1])
# Make first column (#OTU ID) the rownames of the new matrix
rownames(otu_mat) <- otu$`#OTU ID`
# Use the "otu_table" function to make an OTU table
OTU <- otu_table(otu_mat, taxa_are_rows = TRUE) 
class(OTU)
```
#### Format sample metadata ####
```{r}
# Save everything except sampleid as new data frame
samp_df <- as.data.frame(meta[,-1])
# Make sampleids the rownames
rownames(samp_df)<- meta$'sample-id'
# Make phyloseq sample data with sample_data() function
SAMP <- sample_data(samp_df)
class(SAMP)
```

#### Formatting taxonomy ####

```{r}
# Convert taxon strings to a table with separate taxa rank columns
tax_mat <- tax %>% select(-Confidence)%>%
  separate(col=Taxon, sep="; "
           , into = c("Domain","Phylum","Class","Order","Family","Genus","Species")) %>%
  as.matrix() # Saving as a matrix
# Save everything except feature IDs 
tax_mat <- tax_mat[,-1]
# Make sampleids the rownames
rownames(tax_mat) <- tax$`Feature ID`
# Make taxa table
TAX <- tax_table(tax_mat)
class(TAX)
```


#### Create phyloseq object ####

```{r}
# Merge all into a phyloseq object
hiv <- phyloseq(OTU, SAMP, TAX, phylotree)
```

#### Looking at phyloseq object #####

```{r}
# View components of phyloseq object with the following commands
otu_table(hiv)
sample_data(hiv)
tax_table(hiv)
phy_tree(hiv)
```

######### ANALYZE ##########
```{r}
# Remove non-bacterial sequences, if any
hiv_filt <- subset_taxa(hiv,  Domain == "d__Bacteria" & Class!="c__Chloroplast" & Family !="f__Mitochondria")
# Remove ASVs that have less than 5 counts total
hiv_filt_nolow <- filter_taxa(hiv_filt, function(x) sum(x)>5, prune = TRUE)
# Remove samples with less than 100 reads
hiv_filt_nolow_samps <- prune_samples(sample_sums(hiv_filt_nolow)>100, hiv_filt_nolow)
# Remove samples where month is na
hiv_final <- subset_samples(hiv_filt_nolow_samps, !is.na(month) )
```

```{r}
# Rarefy samples
# rngseed sets a random number. If you want to reproduce this exact analysis, you need
# to set rngseed the same number each time
# t transposes the table to use rarecurve function
# cex decreases font size
rarecurve(t(as.data.frame(otu_table(hiv_final))), cex=0.1)
hiv_rare <- rarefy_even_depth(hiv_final, rngseed = 1, sample.size = 12082)
```

##### Saving #####
```{r}
save(hiv_final, file="hiv_final.RData")
save(hiv_rare, file="hiv_rare.RData")
```